---
title: "apoptosisQuantification: Assessment of apoptosis and necroptosis using proteomics"
author:
 - name: Thomas Naake
   mail: thomas.naake@embl.de, thomasnaake@googlemail.com
   affiliation: European Molecular Biology Laboratory, Meyerhofstrasse 1, 69117 Heidelberg
package: apoptosisQuantification
      
output:  
    BiocStyle::html_document:
        toc_float: true
bibliography: apoptosisQuantification-citations.bib
vignette: >
    %\VignetteIndexEntry{Assessment of apoptosis and necroptosis using proteomics"}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteKeywords{Mass Spectrometry, MS, Proteomics, QC, QA}
    %\VignettePackage{apoptosisQuantification-vignette}
    %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r env, include=FALSE, echo=FALSE, cache=FALSE}
library("knitr")
opts_chunk$set(stop_on_error = 1L)
suppressPackageStartupMessages(library("apoptosisQuantification"))
```

# Questions and bugs {-}

`apoptosisQuantification` is currently under active development. If you
discover any bugs, typos or develop ideas of improving
`apoptosisQuantification` feel free to raise an issue via
[GitHub](https://github.com/tnaake/apoptosisQuantification) or
send a mail to the developer.

# Prepare the environment and load the data {#sec-prepare}
To install `apoptosisQuantification` enter the following to the `R` console

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("devtools", quietly = TRUE))
    BiocManager::install("devtools")
devtools::install("tnaake/apoptosisQuantification")
```

# Introduction {#sec-intro}

When performing cell line experiments it is of outermost interest to capture
variation originating from the treatment, while controlling external 
effects. One such effect is the onset of apoptosis that results in cell death
and might overlay the biological effects of interest.

We present here the `apoptosisQuantification` package that allows to quantify
the level and effects of apoptosis (and necroptosis). We use here a list of
markers that are deregulated in apoptosis and necroptosis.

Classical necrotic markers include lactate deydrogenase (LDH),high-mobility 
group B1 (HMGB1), myoglobin, enolase, and 14-3-3 proteins. Commonly, a 
protein release is observed during apoptosis and necroptosis for human
myeloid/lymphoma cell lines and human primary macrophage cells as
observed time course experiments (cf. @Tanzer2020). The work of @Tanzer2020
looked at differential expression of proteins during a time-course experiments
where histiocytic lympoma cell lines (U937) were induced to apoptosis 
(TNF+SM treatment, 
SM: birinapant, TNF: tumor necrosis factor) 
and necroptosis (TNF+SM+Idun-6556, Idun-6556: caspase inhibitor IDUN-6556).

# Load marker proteins

The `readMarkers` function loads a list of apoptosis or necroptosis markers 
taken from the time-course experiment of @Tanzer2020.

We first load a `tibble` that stores the marker proteins together with the 
observed maximum absolute fold change for significant time points and the 
number of time points when the difference between treated and control cell lines
were significant.

```{r read_markers}
markers_apoptosis <- readMarkers(type = "apoptosis", fc = 2, n = 1)
markers_necroptosis <- readMarkers(type = "necroptosis", fc = 2, n = 1)
```

The returned `tibble` contains three columns:

 - `protein`: name of the marker, 
 - `fold_change`: maximum/minimum fold change from the time-course experiment 
    (absolute value is taken to define the maximum change), 
 - `significant_n`: number of significant time-points.


# Assessing apoptosis and necroptosis via loadings

This approach looks at the distribution of the loadings (from PCA) from markers
and non-markers. If there are effects between the different samples caused
by apoptosis/necroptosis we expect that the distribution of marker loadings
is higher than the distribution of non-marker loadings.

```{r}

```

# Assessing apoptosis and necroptosis via mitochondrial protein quantification

# Assessing apoptosis and necroptosis via decoupleR



# Appendix {-}

## Session information {-}

All software and respective versions to build this vignette are listed here:

```{r session,eval=TRUE, echo=FALSE}
sessionInfo()
```

## References

